---
title: "forest-fires-in-portugal"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Load packages
```{r load-pacakges}
library("readr")
library("dplyr")
library("ggplot2")
library("purrr")
```

## Import dataset

```{r import-csv, echo = TRUE}
ff <- read_csv("forestfires.csv")
head(ff)
```

## Questions:
1. During which months are forest fires most common?
2. On which days of the week are forest fires most common?

## 1. Number of forest fires by month
```{r fires-by-month}
fireinmonths <- ff %>%
  group_by(month) %>%
  summarize(count = n())
```

```{r plot-by-month}
ggplot(fireinmonths) +
  aes(x = month, y = count) + 
  geom_bar(stat = "identity") + 
  labs(title = "Number of fires occurred in each month", x = "Month", y = "Number fires") + 
  theme(panel.background = element_rect(fill = "white"))
```
## 2. Number of forest fires by a week-day
```{r fires-per-weekday}
fireindays <- ff %>%
  group_by(day) %>%
  summarize(count_weekday = n())
```

```{r plot-by-weekday}
ggplot(fireindays) +
  aes(x = day, y = count_weekday) + 
  geom_bar(stat = "identity") + 
  labs(title = "Number of fires occurred by each day", x = "Day of the week", y = "Number of fires") + 
  theme(panel.background = element_rect(fill = "white"))
```

## Re-ordering months for easier readability
```{r add-factor}
ff <- ff %>%
  mutate(month = factor(month, c("jan", "feb", "mar", "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec"))) %>%
  mutate(day = factor(day, c("mon", "tue", "wed", "thu", "fri", "sat", "sun")))

```

## Plots after using factor

## Month-wise
```{r fires-per-month}
fireinmonths <- ff %>%
  group_by(month) %>%
  summarize(count = n())
```

```{r plot-again-by-month}
ggplot(fireinmonths) +
  aes(x = month, y = count) + 
  geom_bar(stat = "identity") + 
  labs(title = "Number of fires occurred in each month", x = "Month", y = "Number fires") + 
  theme(panel.background = element_rect(fill = "white"))
```

## Weekday-wise
```{r fires-per-weekday-again}
fireindays <- ff %>%
  group_by(day) %>%
  summarize(count_weekday = n())
```


```{r plot-again}
ggplot(fireindays) +
  aes(x = day, y = count_weekday) + 
  geom_bar(stat = "identity") + 
  labs(title = "Number of fires occurred by each day", x = "Day of the week", y = "Number of s fires") + 
  theme(panel.background = element_rect(fill = "white"))

```
## What are the causes?
 
 Let's explore the different variables like:
 FFMC
 DMC - 
 DC - Drought code
 ISI - Initial spread index
 temp - Temperature
 RH - Relative Humidity
 wind - Wind
 rain - Rain
 
## Box plots by month
```{r function-for-multiple-plots}
create_box_by_month <- function(x, y) {
  ggplot(ff) +
    aes_string(x, y) + 
    geom_boxplot()
}

x_var <- c("month")
y_var <- c("FFMC", "DMC","DC", "ISI", "temp", "RH", "wind", "rain")
  
map2(x_var, y_var, create_box_by_month)
```
#Box Plots by week day
```{r function-for-multiple-plots-by-day}
create_box_by_day <- function(x, y) {
  ggplot(ff) +
    aes_string(x, y) + 
    geom_boxplot()
}

X_var <- c("day")
Y_var <- c("FFMC", "DMC","DC", "ISI", "temp", "RH", "wind", "rain")
  
map2(X_var, Y_var, create_box_by_day)

```
## Which variables are related to forest fire severity?
Area is the variable which helps us measure the fire severity.
```{r}
create_scatter <- function(x,y){
  ggplot(data = ff) +
    aes_string(x,y) +
    geom_point()
}

Y_scat <- c("area")
X_scat <- c("FFMC", "DMC","DC", "ISI", "temp", "RH", "wind", "rain")

map2(X_scat, Y_scat, create_scatter)
```
As most of the area points are in the lower part of the plots, let's filter out data and plot the same scatter plots.
```{r}
ggplot(ff)+
  aes(area) +
  geom_histogram(bins = 20)
```
The above histogram tells that the burnt area concentration is below 500 ha per entry. 
```{r}
area_is_zero <- ff %>%
  filter(area == 0)
area_without_outliers <- ff %>%
  filter(area > 0 & area < 400)
area_0_and_50 <- ff %>%
  filter(area > 0 & area <= 50)

```

```{r}
create_scatter <- function(x,y){
  ggplot(data = area_0_and_50) +
    aes_string(x,y) +
    geom_point()
}

Y_scat <- c("area")
X_scat <- c("FFMC", "DMC","DC", "ISI", "temp", "RH", "wind", "rain")

map2(X_scat, Y_scat, create_scatter)

```
With no clear patterns after plotting without outliers and removing zero area values separately, plotting for the area in the range [0,50] shows `FFMC` has a positive trend with `area`. The plot of rain and area shows that there was no rain when any amount of area occurred, which is a trivial insight. No other variables show any noticeable relation with area.
